---
title: "Normalization"
output: html_document
---

# Pamgene: Testing different normalization methods

## Load packages 

```{r include=FALSE}
library(tidyverse)
library(plotly)
library(lumi)
library(reshape2)
```

## Read in the data

```{r include=FALSE}
raw <- read.csv("C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/cc_cpe_ets.txt", skip = 1)
```

```{r include=FALSE}
str(raw)
```

```{r include=FALSE}
raw['ArrayID'] <- paste0(raw$Barcode, "_", raw$Array)
```

Visualize raw data. 

```{r echo=FALSE}
ggplot(raw, aes(ETS_Identity, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
```


### Transform data

```{r include=FALSE}
# take generalized logarithm (test)
raw['glog'] <- log2(raw$ETS_Identity + sqrt((raw$ETS_Identity*raw$ETS_Identity +1)))
# remove negative values
raw$ETS_Identity[raw$ETS_Identity < 0] <- 0
# take logarithm
raw['log'] <- log2(raw$ETS_Identity+1)
```

Log of all data

```{r echo=FALSE}
p <-  ggplot(raw, aes(log, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

## Normalize REF samples CC study

### Select data

```{r include=FALSE}
#Adapt input.mat function
input.mat = function(df, vcol = "ETS_Identity", ncol = 9){
  wide <- spread(df, ID, value = vcol)
  mat <- wide[,-c(1:ncol)] # probes in columns
  rownames(mat) <- wide$ArrayID
  anno <- wide[,c(1:ncol)]
  result = list(X = as.matrix(mat), An = anno)
}
```


```{r include=FALSE}
cc.ref.raw = raw %>%
  select(-one_of('log', 'glog')) %>% 
  filter(Study == "METHOD-CC") %>%
  filter(Grouping == "REF") %>% 
  input.mat()

dim(cc.ref.raw$X)
class(cc.ref.raw$X)
dim(cc.ref.raw$An)
class(cc.ref.raw$An)
```

New since R.4.0 => matrix always gets class array as well from now on (be careful for this analysis some older R functions struggle with this)

```{r include=FALSE}
#Remove negative values
cc.ref.mat.nn <-  cc.ref.raw$X
cc.ref.mat.nn[cc.ref.mat.nn<0] <- 0

#Take logarithm, offset = 1
cc.ref.mat.log <- log2(cc.ref.mat.nn +1)
#cc.ref.mat.glog <- log2(cc.ref.raw$X + sqrt((cc.ref.raw$X*cc.ref.raw$X +1)))
```

Log of reference data CC study

```{r echo=FALSE}
p <-  ggplot(melt(cc.ref.mat.log), aes(value, group = Var1)) +
  stat_density(geom = "line", position = "identity")
fig <- ggplotly(p)
fig
```

### Robust spline

Robust splines on logarithmized data

The robust spline normalization (RSN) algorithm combines the features of quantile and lowess normalization. Shoulds be applied on variance-stabilized data (e.g. vsn/log data).

```{r include=FALSE}
rsn <- lumi::rsn(t(cc.ref.mat.log))
```

```{r}
rsn_long <-  reshape2::melt(rsn)
colnames(rsn_long) <-  c('Probe', 'ArrayID', 'value_rsn')
rsn_long <- left_join(rsn_long, cc.ref.raw$An, by = 'ArrayID')
```

```{r}
p <- ggplot(rsn_long, aes(value_rsn, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

### Vsn

#### Ref CC

Classic vsn on reference data CC study

```{r include=FALSE}
vsn::meanSdPlot(t(cc.ref.raw$X))
dvsn <- vsn::vsn2(t(cc.ref.raw$X)) 
vsn::meanSdPlot(dvsn)
```

```{r include=FALSE}
dvsnm <- dvsn@hx
```

```{r include=FALSE}
dvsn_long <-  reshape2::melt(dvsnm)
colnames(dvsn_long) <-  c('Probe', 'ArrayID', 'value')
dvsn_long <- left_join(dvsn_long, cc.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsn_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```
Outlier array: 650001519_A4

#### Ref CC No Outlier

Repeat procedure without outlier array

```{r include=FALSE}
cc.ref.raw.no <- cc.ref.raw$X[!rownames(cc.ref.raw$X) %in% c('650001519_A4'), ]
```

```{r include=FALSE}
dvsno <- vsn::vsn2(t(cc.ref.raw.no)) 
vsn::meanSdPlot(dvsno)
```

```{r include=FALSE}
dvsnom <- dvsno@hx
```

```{r include=FALSE}
dvsno_long <-  reshape2::melt(dvsnom)
colnames(dvsno_long) <-  c('Probe', 'ArrayID', 'value')
dvsno_long <- left_join(dvsno_long, cc.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsno_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

#### Ref CPE

```{r include=FALSE}
cpe.ref.raw = raw %>%
  select(-one_of('log', 'glog')) %>% 
  filter(Study == "METHOD-CPE") %>%
  filter(Grouping == "REF") %>% 
  input.mat()

dim(cpe.ref.raw$X)
class(cpe.ref.raw$X)
dim(cpe.ref.raw$An)
class(cpe.ref.raw$An)
```

```{r include=FALSE}
dvsn <- vsn::vsn2(t(cpe.ref.raw$X)) 
vsn::meanSdPlot(dvsn)
```

```{r include=FALSE}
dvsnm <- dvsn@hx
```

```{r include=FALSE}
dvsn_long <-  reshape2::melt(dvsnm)
colnames(dvsn_long) <-  c('Probe', 'ArrayID', 'value')
dvsn_long <- left_join(dvsn_long, cpe.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsn_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```
Outlier array: 650055110_A2

```{r include=FALSE}
cpe.ref.raw.no <- cpe.ref.raw$X[!rownames(cpe.ref.raw$X) %in% c('650055110_A2'), ]
```

```{r include=FALSE}
dvsno_cpe <- vsn::vsn2(t(cpe.ref.raw.no)) 
vsn::meanSdPlot(dvsno_cpe)
```

```{r include=FALSE}
dvsnom_cpe <- dvsno_cpe@hx
```

```{r include=FALSE}
dvsno_cpe_long <-  reshape2::melt(dvsnom_cpe)
colnames(dvsno_cpe_long) <-  c('Probe', 'ArrayID', 'value')
dvsno_cpe_long <- left_join(dvsno_cpe_long, cpe.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsno_cpe_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

Conclusion both for the CC Ref and the CPE Ref one 'abbarent'outlier' array.
Extra filtering steps?


### Classic normalisation procedure without outliers
#### Towards CC

```{r include=FALSE}
#remove outliers
all.raw = raw %>%
  filter(!(ArrayID %in% c('650001519_A4', '650055110_A2'))) %>% 
  select(-one_of('log', 'glog')) %>% 
  input.mat()

dim(all.raw$X)
dim(all.raw$An)
```

```{r include=FALSE}
#like classic normalisation procedure used in pipeline, without outliers
submat_norm_ref <- vsn::vsn2(t(all.raw$X), reference = dvsno)
vsn::meanSdPlot(submat_norm_ref)
```

```{r include=FALSE}
nccr_long <-  reshape2::melt(submat_norm_ref@hx)
colnames(nccr_long) <-  c('Probe', 'ArrayID', 'value')
nccr_long <- left_join(nccr_long, all.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(nccr_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

```{r}
write.csv(nccr_long,"C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/classic_svn_towards_cc_nooutliers.csv")
```

Adapt lts.quantile parameter (most conservative option)

```{r include=FALSE}
submat_norm_ref <- vsn::vsn2(t(all.raw$X), reference = dvsno, lts.quantile=0.5)
vsn::meanSdPlot(submat_norm_ref)
```

```{r include=FALSE}
dvsn_long <-  reshape2::melt(submat_norm_ref@hx)
colnames(dvsn_long) <-  c('Probe', 'ArrayID', 'value')
dvsn_long <- left_join(dvsn_long, all.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsn_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

<!-- ```{r eval=FALSE, include=FALSE} -->
<!-- write.csv(dvsn_long,"C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/classic_svn_towards_cc_nooutliers_ltsrobust.csv") -->
<!-- ``` -->

#### Towards CPE

Repeat classic normalisation procedure but normalize towards reference samples CPE

```{r include=FALSE}
norm_cpe_ref <- vsn::vsn2(t(all.raw$X), reference = dvsno_cpe)
vsn::meanSdPlot(norm_cpe_ref)
```

```{r include=FALSE}
ncper_long <-  reshape2::melt(norm_cpe_ref@hx)
colnames(ncper_long) <-  c('Probe', 'ArrayID', 'value')
ncper_long <- left_join(ncper_long, all.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(ncper_long, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

```{r}
write.csv(ncper_long,"C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/classic_svn_towards_cpe_nooutliers.csv")
```

### vsn without affine calibration 

#### CC

```{r include=FALSE}
dvsnc <- vsn::vsn2(t(cc.ref.raw$X), calib= 'none')
dvsnc <- dvsnc@hx
dvsn_longc <-  reshape2::melt(dvsnc)
colnames(dvsn_longc) <-  c('Probe', 'ArrayID', 'value')
dvsn_longc <- left_join(dvsn_longc, cc.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(dvsn_longc, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

#### Combine with robust splines

```{r include=FALSE}
vsnc_rsn <- lumi::rsn(dvsnc)
vsnc_rsnl <-  reshape2::melt(vsnc_rsn)
colnames(vsnc_rsnl) <-  c('Probe', 'ArrayID', 'value')
vsnc_rsnl <- left_join(vsnc_rsnl, cc.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(vsnc_rsnl, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```
Outlier array seen before seems to dissapear


#### CPE

```{r include=FALSE}
n <- vsn::vsn2(t(cpe.ref.raw$X), calib = 'none')
n <- n@hx
nl <-  reshape2::melt(n)
colnames(nl) <-  c('Probe', 'ArrayID', 'value')
nl <- left_join(nl, cpe.ref.raw$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(nl, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour=RunID))
fig <- ggplotly(p)
fig
```

#### VSN without affine + RSN All data

```{r include=FALSE}
cc.ref.raw.no <- cc.ref.raw$X[!rownames(cc.ref.raw$X) %in% c('650001519_A4'), ]
```

```{r include=FALSE}
dvsnc <- vsn::vsn2(t(cc.ref.raw.no), calib= 'none')
```

```{r include=FALSE}
norm_vsn_nc <- vsn::vsn2(t(all.raw$X), calib = 'none')
vsn::meanSdPlot(norm_vsn_nc)
```

```{r include=FALSE}
#target array: median normalized array
norm_vsn_rsn <- lumi::rsn(norm_vsn_nc@hx, targetArray = rowMedians(dvsnc@hx))
norm_vsn_rsnl <-  reshape2::melt(norm_vsn_rsn)
colnames(norm_vsn_rsnl) <-  c('Probe', 'ArrayID', 'value')
norm_vsn_rsnl <- left_join(norm_vsn_rsnl, all.raw$An, by = 'ArrayID')
```

target array, vsn normalization can be further adapted

```{r echo=FALSE}
p <- ggplot(norm_vsn_rsnl, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

```{r}
write.csv(norm_vsn_rsnl,"C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/rsn_svn_towards_cc_nooutliers.csv")
```


### Rank invariant


```{r}

#Adapted function from HTqPCR package 
library(HTqPCR)
library(affy)

# Macx is now set on log data, can be adapted => solution saturation?

rin <- function (data, norm = "norm.rankinvariant", 
    rank.type = "pseudo.median", max = 20, 
    verbose = TRUE)
{
    data.norm <- data
    method <- match.arg(norm, c("quantile", "scale.rankinvariant",
        "norm.rankinvariant"))
    if (method %in% c("scale.rankinvariant", "norm.rankinvariant")) {
        index <- data > max
        data.max <- data
        data.max[index] <- NA
        if (rank.type == "pseudo.median") {
            ref.data <- apply(data.max, 1, median, na.rm = TRUE)
        }
        else if (rank.type == "pseudo.mean") {
            ref.data <- apply(data.max, 1, mean, na.rm = TRUE)
        }
        na.index <- is.na(ref.data)
        ref.data[na.index] <- 20
        data.rankinvar <- apply(data, 2, normalize.invariantset,
            ref = ref.data)
    }
    switch(method, quantile = {
        data.norm <- normalizeQuantiles(data)
    }, scale.rankinvariant = {
        ri.genes <- sapply(data.rankinvar, "[[", "i.set")
        ri.genes[index] <- FALSE
        ri.genes[na.index, ] <- FALSE
        ri.count <- rowSums(ri.genes)
        if (missing(scale.rank.samples)) scale.rank.samples <- ncol(data) -
            1
        ri.index <- ri.count >= scale.rank.samples
        if (sum(ri.index) == 0) stop(paste("No rank invariant probes were found across",
            scale.rank.samples, "samples"))
        ri.mean <- colMeans(data[ri.index, , drop = FALSE])
        ri.scale <- ri.mean/ri.mean[1]
        data.norm <- t(t(data) * ri.scale)
        if (verbose) {
            cat(c("Scaling Ct values\n\tUsing rank invariant probes:",
                paste(featureNames(q)[ri.index], collapse = " "),
                "\n"))
            cat(c("\tScaling factors:", format(ri.scale,
                digits = 3), "\n"))
        }
    }, norm.rankinvariant = {
        if (verbose) cat("Normalizing Ct values\n\tUsing rank invariant probes:\n")
        for (i in 1:ncol(data)) {
            ri.sub <- data.rankinvar[[i]]
            ri.genes <- ri.sub[["i.set"]]
            ri.genes[index[, i]] <- FALSE
            ri.genes[na.index] <- FALSE
            if (sum(ri.genes) == 0) {
                warning(paste("\tNo rank invariant genes were found for sample ",
                  colnames(data)[i], "; sample not normalized\n",
                  sep = ""))
                next
            }
            if (verbose) cat(paste("\t", colnames(data)[i],
                ": ", sum(ri.genes), " rank invariant probes\n",
                sep = ""))
            data.norm[, i] <- as.numeric(approx(ri.sub$n.curve$y,
                ri.sub$n.curve$x, xout = data[, i], rule = 2)$y)
        }
    })

    data.norm
}
```

#### Reference CC 
```{r include=FALSE}
cc.raw.log = raw %>%
  filter(!(ArrayID %in% c('650001519_A4', '650055110_A2'))) %>%
  select(-one_of('ETS_Identity', 'glog')) %>%
  filter(Study == "METHOD-CC") %>%
  filter(Grouping == "REF") %>% 
  input.mat(vcol = "log")
```

```{r}
norm <- rin(t(cc.raw.log$X))
```
All reference samples => all probes should be invariant
Interesting to see which probes do vary (same as probes that are deleted later in pipieline?)
Effect removing/ not accounting these probes before norm?
What can be further adapted => reference is now median over all arrays!! (adapt to only reference data)

#### All data
```{r include=FALSE}
all.raw.log = raw %>%
  filter(!(ArrayID %in% c('650001519_A4', '650055110_A2'))) %>%
  select(-one_of('ETS_Identity', 'glog')) %>%
  input.mat(vcol = "log")
```

```{r}
norm <- rin(t(all.raw.log$X))
```

```{r include=FALSE}
norm_inv_l <-  reshape2::melt(norm)
colnames(norm_inv_l) <-  c('Probe', 'ArrayID', 'value')
norm_inv_l <- left_join(norm_inv_l, all.raw.log$An, by = 'ArrayID')
```

```{r echo=FALSE}
p <- ggplot(norm_inv_l, aes(value, group = ArrayID)) +
  stat_density(geom = "line", position = "identity", aes(colour = RunID))
fig <- ggplotly(p)
fig
```

```{r}
write.csv(norm_inv_l,"C:/Users/laure_vansteenbergen/Liz10_3_PamGene/data/rankinvariant_log_towards_median_nooutliers.csv")
```

